---
title: "ChIPprofileSimilarity"
output:
  html_document:
    theme: null
    toc: yes
  pdf_document:
    theme: null
    toc: yes
---

<!-- rmarkdown v1 -->
<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{ChIPprofileSimilarity: similarity between two ChIP profiles}
-->


Metrics which estimate similarity between two ChIP profiles
====================================================================

Astrid Louise Deschenes, Elsa Bernatchez, Charles Joly Beauparlant, Fabien Claude Lamaze, Rawane Samb, Pascal Belleau and Arnaud Droit.

This package and the underlying ChIPprofileSimilarity code are distributed under the Artistic license 2.0. You are free to use and redistribute this software. 


## Introduction

The **ChIPprofileSimilarity** package calculates metrics to estimate the level of similarity between two ChIP profiles.

The four metrics are:

* **RATIO_AREA**: The ratio between the profile areas. The larger value is always divided by the smaller value. `NA` if minimal threshold is not respected.

* **DIFF_POS_MAX**: The difference between the maximal peaks positions. The difference is always a positive value. A profile can have more than one position with the maximum value. In that case, the median position is used. A threshold argument can be set to consider all positions within a certain range of the maximum value.

* **RATIO_MAX_MAX**: The ratio between the maximal peaks values in each profile. The larger value is always divided by the smaller value. `NA` if minimal threshold is not respected.

* **RATIO_INTERSECT**: The ratio between the intersection area and the total area of the two profiles. `NA` if minimal threshold is not respected.
   

## Loading ChIPprofileSimilarity package

```{r libraryLoad}
suppressMessages(library(ChIPprofileSimilarity))
```
 
 
## Inputs

### ChIP profiles vectors

To estimate the level of similarity between two ChIP profiles for a specific region, two `vector` 
containing the RPM values of the two ChIP profiles for each position of the selected region. 
Both `vector` should have the same length and should not contain any negative value.

### Threshold values

When the metric includes a ratio calculation, a threshold for the denominator is assigned to the metric. 
The threshold has to be positive numerical value. When the threshold is not respected, the metric is
not calculated and the `NA` value is assigned to it.

The three threshold are:

* **ratioAreaThreshold**: The minimum denominator accepted to calculate the ratio of the area between 
both profiles. Default = 1.

* **ratioMaxMaxThreshold**: The minimum denominator accepted to calculate the ratio of the maximal peaks values between both profiles. Default = 1.

* **ratioIntersectThreshold**: The minimum denominator accepted to calculate the ratio of the intersection area of both profiles over the total area. Default = 1.

* **diffPosMaxThresholdMinValue**: The smaller maximum value accepted to calculate the metric. When this value is not respected, it is assumed that no preak is present in the profile. Default = 1.

* **diffPosMaxThresholdMaxDiff**: The larger distance accepted between 2 maximum positions in one profile to calculate the metric. When this distance is not respected, it is assumed that more than one peak is present in the profile. Default=100.

* **diffPosMaxTolerancePercent**: The maximum percentage of variation accepted on the maximum value to consider a position as a included in the maximum positions list. Default=1.

## Metrics calculation

The metrics are calculated using the `similarity` function which takes as arguments the two ChIP profiles vectors and the threshold values.

```{r metricCalculation}
profile1<-c(3,59,6,24,65,112,144,77,53,22,21,12,11)
profile2<-c(15,9,46,44,59,39,27,34,34,4,3,4,2)
metrics<-ChIPprofileSimilarity:::similarity(profile1, profile2, ratioAreaThreshold=5, ratioMaxMaxThreshold=5, ratioIntersectThreshold=12)
```

The `similarity` function returns a list which contains the information about both ChIP-Seq profiles and a list of all four metrics. 

```{r metricReturn}
metrics
```